---
title: "arcstats"
output:
  github_document:
    toc: true
    toc_depth: 3
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo = TRUE, 
  cache = FALSE, 
  message = FALSE
)
```

<!-- badges: start -->
<!-- badges: end -->

The `arcstats` package allows to generate summaries of the minute-level physical activity (PA) data. The default parameters are chosen for the Actigraph activity counts collected with a wrist-worn device; however, the package can be used for all minute-level PA data with the corresponding timepstamps vector.

Below, we demonstrate the use of `arcstats` with the attached, exemplary minute-level Actigraph PA counts data.

## Installation

You can install the released version of arcstats from [GitHub](https://github.com/). Note you may need to install `devtools` package if not yet installed (the line commened below). 


``` r
# install.packages("devtools")
devtools::install_github("martakarass/arcstats")
```

## Documentation 

A PDF with detailed documentation of all methods can be accessed [here](https://www.dropbox.com/s/cfnud6zxntub7a7/arcstats_0.1.1.pdf?dl=1). 

# Using `arcstats` package to compute physical activity summaries

### Reading PA data

Four CSV data sets with minute-level activity counts data are attached to the `arcstats` package. The data file names are stored in `extdata_fnames`.

```{r}
library(arcstats)
library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)

## Read one of the data sets
fpath <- system.file("extdata", extdata_fnames[1], package = "arcstats")
dat   <- as.data.frame(fread(fpath))
rbind(head(dat, 3), tail(dat, 3))
```

The data columns are: 

- `Axis1` - sensor's X axis minute-level counts data,
- `Axis2` - sensor's Y axis minute-level counts data,
- `Axis3` - sensor's Z axis minute-level counts data,
- `vectormagnitude` - minute-level counts data defined as `sqrt(Axis1^2 + Axis2^2 + Axis3^2)`,
- `timestamp` - time-stamps corresponding to minute-level measures. 

```{r, fig.width=8, fig.height=3.5}
## Plot activity counts
ggplot(dat, aes(x = ymd_hms(timestamp), y = vectormagnitude)) + 
  geom_line(size = 0.3, alpha = 0.8) + 
  labs(x = "Time", y = "Activity counts") + 
  theme_gray(base_size = 10) + 
  scale_x_datetime(date_breaks = "1 day", date_labels = "%b %d")
```

### Computing summaries with `activity_stats` method

```{r, results='hide'}
acc    <- dat$vectormagnitude
acc_ts <- ymd_hms(dat$timestamp)

activity_stats(acc, acc_ts)
```

```{r, echo = FALSE}
library(knitr)
library(kableExtra)
tbl_styl_f <- function(x) kable_styling(x, font_size = 14)
```

```{r, echo = FALSE}
out <- activity_stats(acc, acc_ts)
# tbl_1 <- kable(out[, 1:6],   table.attr = 'class="myTable"') %>% tbl_styl_f(.)
tbl_1 <- kable(out[, 1:6]) %>% tbl_styl_f(.)
tbl_2 <- kable(out[, 7:10]) %>% tbl_styl_f(.)
tbl_3 <- kable(out[, 11:14]) %>% tbl_styl_f(.)
tbl_1; tbl_2; tbl_3
```

### Output explained 

To explain `activity_stats` method output, we define activity count, active/non-active minute, and active/non-active bout. 

- Activity count (AC) - a munute-level metric of PA volume
- Active minute - AC labeled as active (non-sedentary). For Wrist-worn Actigraph
we use AC>=1853 (defaut threshold).
- Active bout - sequence of >=1 consecutive active minute(s). 
- Valid day - a day with less than 10% of the non-wear time (see `?get_wear_flag` for wear/non-wear time detection details). 

Meta information: 

- `n_days` - number of days (unique day dates) of data collection.
- `n_valid_days` - number of days (unique day dates) of data collection determined as valid days. 
- `wear_time_on_valid_days` - average number of wear-time minutes across valid days.

Summaries of PA volumes metrics: 

- `tac` - TAC, Total activity counts per day - sum of AC measured on valid days divided by the number of valid days.
- `tlac` - TLAC, Total-log activity counts per day  - sum of log(1+AC) measured on valid days divided by the number of valid days. Here 'log' denotes the natural logarithm.
- `ltac` - LTAC, Log-total activity counts - natural logarithm of TAC.
- `time_spent_active` - Average number of active minutes per valid day.
- `time_spent_nonactive` - Average number of sedentary minutes per valid day.

Summaries of PA fragmentation metrics: 

- `astp` - ASTP, active to sedentary transition probability on valid days. 
- `satp` - SATP, sedentary to active transition probability on valid days. 
- `no_of_active_bouts` - Average number of active minutes per valid day.
- `no_of_nonactive_bouts` - Average number of sedentary minutes per valid day.
- `mean_active_bout` - Average duration (in minutes) of an active bout on valid days.
- `mean_nonactive_bout` - Average duration (in minutes) of a sedentary bout on valid days.


# Additional`activity_stats` method options

### Summarizing PA only in a chosen time-of-day subset 

The `subset_minutes` argument allows to specify subset of a day's minutes where activity summaries should be computed. There are 1440 minutes in a 24-hour day where `1` denotes 1st minute of the day (from 00:00 to 00:01), and `1440` denotes the last minute (from 23:59 to 00:00). 

Here, we summarize PA observed between 12:00 AM and 6:00 AM.

```{r, results = 'hide'}
subset_12am_6am <- 1 : (6 * 1440/24)
activity_stats(acc, acc_ts, subset_minutes = subset_12am_6am) 
```

```{r, echo = FALSE}
out <- activity_stats(acc, acc_ts, subset_minutes = subset_12am_6am) 
tbl_1 <- kable(out[, 1:6]) %>% tbl_styl_f(.)
tbl_2 <- kable(out[, 7:10]) %>% tbl_styl_f(.)
tbl_3 <- kable(out[, 11:14]) %>% tbl_styl_f(.)
tbl_1; tbl_2; tbl_3
```

By default, column names have a suffix added to denote that a subset of munutes was used (here, `_0to6only`). This can be disabled by setting `adjust_out_colnames` to `FALSE`. 

```{r, results = 'hide'}
subset_12am_6am = 1 : (6/24 * 1440)
subset_6am_12pm = (6/24 * 1440 + 1) : (12/24 * 1440) 
subset_12pm_6pm = (12/24 * 1440 + 1) : (18/24 * 1440) 
subset_6pm_12am = (18/24 * 1440 + 1) : (24/24 * 1440) 
out <- rbind(
  activity_stats(acc, acc_ts, subset_minutes = subset_12am_6am, adjust_out_colnames = FALSE),
  activity_stats(acc, acc_ts, subset_minutes = subset_6am_12pm, adjust_out_colnames = FALSE),
  activity_stats(acc, acc_ts, subset_minutes = subset_12pm_6pm, adjust_out_colnames = FALSE),
  activity_stats(acc, acc_ts, subset_minutes = subset_6pm_12am, adjust_out_colnames = FALSE))
rownames(out) <- c("12am-6am", "6am-12pm", "12pm-6pm", "6pm-12am")
out
```

```{r, echo = FALSE}
tbl_1 <- kable(out[, 1:6]) %>% tbl_styl_f(.)
tbl_2 <- kable(out[, 7:10]) %>% tbl_styl_f(.)
tbl_3 <- kable(out[, 11:14]) %>% tbl_styl_f(.)
tbl_1; tbl_2; tbl_3
```


### Summarizing PA with a chosen time-of-day excluded

The `exclude_minutes` argument allows to specify subset of a day's minutes excluded for computing activity summaries. 

Here, we summarize PA while excluding observatins between 11:00 PM and 5:00 AM. 

```{r, results = 'hide'}
subset_11pm_5am <- c(
  (23 * 1440/24 + 1) : 1440,   ## 11:00 PM - midnight
  1 : (5 * 1440/24)            ## midnight - 5:00 AM
) 
activity_stats(acc, acc_ts, exclude_minutes = subset_11pm_5am) 
```

```{r, echo = FALSE}
out <- activity_stats(acc, acc_ts, exclude_minutes = subset_11pm_5am) 
tbl_1 <- kable(out[, 1:6]) %>% tbl_styl_f(.)
tbl_2 <- kable(out[, 7:10]) %>% tbl_styl_f(.)
tbl_3 <- kable(out[, 11:14]) %>% tbl_styl_f(.)
tbl_1; tbl_2; tbl_3
```


### Summarizing PA excluding in-bed time 

The `in_bed_time` and `out_bed_time` arguments allow to provide day-specific in-bed periods to be excluded from analysis. 

Here, we summarize PA excluding in-bed time estimated by ActiLife software. 

##### ActiLife-estimated in-bed data 

The ActiLife-estimated in-bed data file is attached to the `arcstats` package. The sleep data columns relevant are,

    - `Subject Name` - subject IDs coressponding to AC data, stored in `extdata_fnames`,
    - `In Bed Time` - ActiLife-estimated start of in-bed interval for each day of the measurement, 
    - `Out Bed Time` - ActiLife-estimated end of in-bed interval. </br> 

```{r}
## Read sleep details data file
SleepDetails_fname <- "BatchSleepExportDetails_2020-05-01_14-00-46.csv"
SleepDetails_fpath <- system.file("extdata", SleepDetails_fname, package = "arcstats")
SleepDetails       <- as.data.frame(fread(SleepDetails_fpath))

## Filter sleep details data to keep data correcponding to current counts data file
SleepDetails_sub <-
    SleepDetails %>%
    filter(`Subject Name` == "ID_1") %>%
    mutate(`In Bed Time`  = mdy_hms(`In Bed Time`),
           `Out Bed Time` = mdy_hms(`Out Bed Time`)) %>%
    select(`Subject Name`, `In Bed Time`, `Out Bed Time`) 
SleepDetails_sub
```

Finally, we use in/out-bed time `POSIXct` vectors in `activity_stats` method. 

```{r, results = 'hide'}
in_bed_time  <- SleepDetails_sub[, "In Bed Time"]
out_bed_time <- SleepDetails_sub[, "Out Bed Time"]

activity_stats(acc, acc_ts, in_bed_time = in_bed_time, out_bed_time = out_bed_time) 
```

```{r, echo = FALSE}
out <- activity_stats(acc, acc_ts, exclude_minutes = subset_11pm_5am) 
tbl_1 <- kable(out[, 1:6]) %>% tbl_styl_f(.)
tbl_2 <- kable(out[, 7:10]) %>% tbl_styl_f(.)
tbl_3 <- kable(out[, 11:14]) %>% tbl_styl_f(.)
tbl_1; tbl_2; tbl_3
```



# Components of `activity_stats` method 

The primary method `activity_stats` is composed of several steps implemented in their respective functions. Below, we demonstrate how to produce `activity_stats` results step by step with these functions. 

We reuse the objects: 

- `acc` - numeric vector; minute-level activity counts data,
- `acc_ts` - POSIXct vector; minute-level time of `acc` data collection. 

```{r}
df <- data.frame(acc = acc, acc_ts = acc_ts)
rbind(head(df, 3), tail(df, 3))
```


### Expand the length of minute-level AC vector to integer number of full 24-hour periods by NA-padding with `midnight_to_midnight`

- In the returned vector, the first observation corresponds to minute of `00:00-00:01` on the first day of data collection, and last observation corresponds to minute of `23:50-00:00` on the last day of data collection. 
- Entries corresponding to non-measured minutes are filled with `NA`.

Here, collected data cover total of `7*24*1440 = 10080` minutes (from `2018-07-13 10:00:00` to `2018-07-20 09:59:00`), but spans `8*24*1440 = 11520` minutes of full midnight-to-midnight days (from `2018-07-13 00:00:00` to `2018-07-20 23:59:00`). 

```{r}
acc <- midnight_to_midnight(acc = acc, acc_ts = acc_ts)

## Vector length on non NA-obs, vector length after acc 
c(length(acc[!is.na(acc)]), length(acc))
```


### Get wear/non-wear flag with `get_wear_flag`

Function `get_wear_flag` computes wear/non-wear flag (`1/0`) for each minute of activity counts data. Method implements wear/non-wear detection algorithm proposed by Choi et al. (2011). 

- The returned vector has value `1` for wear and `0` for non-wear flagged minute.
- If there is an `NA` entry in a data imput vector, then the returned vector will have a corresponding entry set to `NA` too.

```{r}
wear_flag <- get_wear_flag(acc)

## Proportion of wear time across the days
wear_flag_mat <- matrix(wear_flag, ncol = 1440, byrow = TRUE)
round(apply(wear_flag_mat, 1, sum, na.rm = TRUE) / 1440, 3)
```


### Get valid/non-valid day flag with `get_valid_day_flag`

Function `get_valid_day_flag` computes valid/non-valid day flag (`1/0`) for each minute of activity counts data. 

Here, 4 out of 8 days have more that 10% (144 minutes) of missing data.

```{r}
valid_day_flag <- get_valid_day_flag(wear_flag)

## Compute number of valid days
valid_day_flag_mat <- matrix(valid_day_flag, ncol = 1440, byrow = TRUE)
apply(valid_day_flag_mat, 1, mean, na.rm = TRUE)
```


### Impute missing data with `impute_missing_data`

Function `impute_missing_data` imputes missing data in valid days based on the "average day profile". 

- An "average day profile" is computed as a minute-wise average of wear-time AC across valid days.

- AC data are imputed from "average day profile" for minutes identified as non-wear in days identified as valid. 

Here, all four valid days have 100% of wear time (recall result of proportion of wear time across all days: `0.583 1.000 0.874 0.679 1.000 1.000 1.000 0.338`). We hence demonstrate the `impute_missing_data` method by artificially replacing 1h (4%) of a valid day with non-wear. 

JU: I vave no idea what the last paragraph means. I would remove it completely.

```{r}
## Copies of original objects for the purpose of demonstration
acc_cpy  <- acc
wear_flag_cpy <- wear_flag

## Artificially replace 1h (4%) of a valid day with non-wear 
repl_idx <- seq(from = 1441, by = 1, length.out = 60)
acc_cpy[repl_idx] <- 0
wear_flag_cpy[repl_idx] <- 0

## Impute data for minutes identified as non-wear in days identified as valid
acc_cpy_imputed <- impute_missing_data(acc_cpy, wear_flag_cpy, valid_day_flag)

## Compare mean activity count on valid days before and after imputation
c(mean(acc_cpy[which(valid_day_flag == 1)]), 
  mean(acc_cpy_imputed[which(valid_day_flag == 1)]))
```


### Create PA characteristics with `summarize_PA`

Finally, method `summarize_PA` computes physical all PA characteristics.

- Same as `activity_stats`, it returns a data frame with physical activity summaries. 
- Same as `activity_stats`, it accepts arguments to compute physical activity summaries 
    - within a fixed subset of minutes (arg. `subset_minutes`),
    - excluding a fixed subset of minutes (arg. `exclude_minutes`),
    - excluding day-specific in-bed intervals (args `in_bed_time`, `out_bed_time`).
  See `?summarize_PA` for more details. 
  
Here, we summarize PA with the default parameters, across all valid days. 

```{r, results = 'hide'}
summarize_PA(acc, acc_ts, wear_flag, valid_day_flag) 
```

```{r, echo = FALSE}
out <- summarize_PA(acc, acc_ts, wear_flag, valid_day_flag) 
tbl_1 <- kable(out[, 1:6]) %>% tbl_styl_f(.)
tbl_2 <- kable(out[, 7:10]) %>% tbl_styl_f(.)
tbl_3 <- kable(out[, 11:14]) %>% tbl_styl_f(.)
tbl_1; tbl_2; tbl_3
```

It returns the same results as the `activity_stats` function: 

```{r, results = 'hide'}
activity_stats(dat$vectormagnitude, ymd_hms(dat$timestamp))
```

```{r, echo = FALSE}
out <- activity_stats(dat$vectormagnitude, ymd_hms(dat$timestamp))
tbl_1 <- kable(out[, 1:6]) %>% tbl_styl_f(.)
tbl_2 <- kable(out[, 7:10]) %>% tbl_styl_f(.)
tbl_3 <- kable(out[, 11:14]) %>% tbl_styl_f(.)
tbl_1; tbl_2; tbl_3
```



